Overcoming multiple sources of misspecification\\in GWAS summary statistics

### 2021-06-02 paper/paper-misspec-supp.tex
save

\begin{figure}[h]
	\centerline{\includegraphics[width=0.95\textwidth]{simu-misN}}
	\caption{Results for the simulations with sample size misspecification. The GWAS sample size is ``true" when providing the true per-variant sample size, ``max'' when providing instead the maximum sample size as a unique value to be used for all variants,  ``imputed'' (see Methods in main text), or ``any'' when the method does not use this information (the case for C+T).}
	\label{fig:simu-misN}
\end{figure}

@@

\begin{figure}[htb]
	\centerline{\includegraphics[width=0.8\textwidth]{simu_hist_info}}
	\caption{Histogram of the imputation INFO scores of the 20,000 variants from chromosome 22 of the UK Biobank data used in the simulations. \label{fig:hist-info}}
\end{figure}

@@

%\begin{figure}[h]
%	\centerline{\includegraphics[width=0.85\textwidth]{pseudoval}}
%	\caption{For one simulation, scores from validation vs.\ scores from pseudo-validation as described in \cite{mak2017polygenic} using either correlations (the default in lassosum) or p-values for computing local false discovery rates.}
%	\label{fig:pseudoval}
%\end{figure}

>>>>

\begin{figure}[p]
	\centerline{\includegraphics[width=0.95\textwidth]{simu-misN}}
	\caption{Results for the simulations with sample size misspecification. The GWAS sample size is ``true" when providing the true per-variant sample size, ``max'' when providing instead the maximum sample size as a unique value to be used for all variants,  ``imputed'' (see Methods in main text), or ``any'' when the method does not use this information (the case for C+T).}
	\label{fig:simu-misN}
\end{figure}

@@

\begin{figure}[p]
	\centerline{\includegraphics[width=0.8\textwidth]{compare-sd}}
	\caption{Standard deviations from ``true'' genotype calls versus from imputed dosages, colored by INFO scores. \label{fig:compare-sd}}
\end{figure}

\begin{figure}[p]
	\centerline{\includegraphics[width=0.8\textwidth]{compare-beta}}
	\caption{GWAS effect sizes from ``true'' genotype calls versus from imputed dosages, colored by INFO scores. \label{fig:compare-beta}}
\end{figure}

\begin{figure}[p]
	\centerline{\includegraphics[width=0.8\textwidth]{compare-beta-se}}
	\caption{Standard errors of GWAS effect sizes from ``true'' genotype calls versus from imputed dosages, colored by INFO scores. \label{fig:compare-beta-se}}
\end{figure}

\begin{figure}[p]
	\centerline{\includegraphics[width=0.8\textwidth]{compare-info}}
	\caption{INFO scores reported in the UK Biobank vs.\ INFO scores recomputed from the subset of 362,307 European individuals used in this paper. \label{fig:compare-info}}
\end{figure}

\begin{figure}[p]
	\centerline{\includegraphics[width=0.95\textwidth]{simu-info}}
	\caption{Results for the simulations using dosage data. The GWAS sample size is ``true" when providing the true per-variant sample size, ``max'' when providing instead the maximum sample size as a unique value to be used for all variants,  ``imputed'' (see Methods in main text), or ``any'' when the method does not use this information (the case for C+T).}
	\label{fig:simu-info}
\end{figure}

\begin{figure}[p]
	\centerline{\includegraphics[width=0.8\textwidth]{simu_hist_info}}
	\caption{Histogram of the imputation INFO scores of the 40,000 variants from chromosome 22 of the UK Biobank data used in the simulations. \label{fig:hist-info}}
\end{figure}

\begin{figure}[p]
	\centerline{\includegraphics[width=0.8\textwidth]{compare-mult-imp}}
	\caption{GWAS effect sizes and Z-scores from imputed dosages versus from multiple imputation (MI, 10 draws used), ``corrected'' and colored by INFO scores. This shows only one tenth of the simulated variants, because MI takes some time to run. \label{fig:compare-mi}}
\end{figure}

@@

\begin{figure}[p]
	\centerline{\includegraphics[width=0.85\textwidth]{pseudoval}}
	\caption{For one simulation, scores from validation vs.\ scores from pseudo-validation as described in \cite{mak2017polygenic} using either correlations (the default in lassosum) or p-values when computing local false discovery rates.}
	\label{fig:pseudoval}
\end{figure}

### 2021-06-02 paper/paper-misspec.tex
save

We design simulations where variants have different GWAS sample sizes, which is often the case when meta-analyzing GWAS from multiple cohorts without the same genome coverage.
Using 20,000 variants from chromosome 22 (Methods), we simulate quantitative phenotypes assuming an heritability of 20\% and 2000 causal variants.
For half of the variants, we use 100\% of 300,000 individuals for GWAS, but use only 80\% for one quarter of them and 60\% for the remaining quarter. 
We then run C+T, LDpred2-inf, LDpred2(-grid), LDpred2-auto, lassosum, lassosum2, PRS-CS and SBayesR by using either the true per-variant GWAS sample size, or the total sample size. 
We do not show results for SBayesR because it always diverged and for PRS-CS because the overlap with the LD reference they provide is too small.
Averaged over 10 simulations, when providing true per-variant GWAS sample sizes, squared correlations (in \%) between the polygenic scores and the simulated phenotypes are of [TODO: COMPLETE] (Figure \ref{fig:simu-misN}).
Note that C+T does not use this sample size information.
For methods that use the per-variant GWAS sample sizes, predictive performance slightly decreases [TODO: REPLACE NUMBERS] from 17.5 to 17.4 for lassosum, from 18.0 to 17.5 for lassosum2, but dramatically decreases for LDpred2 with new values of 11.5 for LDpred2-grid, 5.7 for LDpred2-auto and 6.8 for LDpred2-inf (Figure \ref{fig:simu-misN}).
Therefore, this extreme simulation scenario shows that LDpred2 can be sensitive to GWAS sample size misspecification, whereas lassosum (and lassosum2) seems little affected by this.

@@

@@

@@

Instead of using a regularized version of the correlation matrix $R$ parameterized by $s$, $R_s = (1 - s) R + s I$, we use $R_{\delta} = R + \delta I$, which makes it clearer that lassosum is also using L2-regularization (therefore elastic-net). 
Then, from \cite{mak2017polygenic}, the solution from lassosum can be obtained by iteratively updating 
\[
\beta_j^{(t)} =
\begin{cases}
\text{sign}\left(u_j^{(t)}\right) \left(\left|u_j^{(t)}\right| - \lambda\right) / \left(\widetilde{X}_j^T \widetilde{X}_j + \delta\right) & \text{if } \left|u_j^{(t)}\right| > \lambda ~, \\
0 & \text{otherwise.}
\end{cases}
\]
where 
\[
u_j^{(t)} = r_j - \widetilde{X}_j^T \left( \widetilde{X} \beta^{(t-1)} - \widetilde{X}_j \beta_j^{(t-1)} \right) ~.
\]
Following notations from \cite{prive2020ldpred2} and denoting $\widetilde{X} = \frac{1}{\sqrt{n-1}} C_n G S^{-1}$, where $G$ is the genotype matrix, $C_n$ is the centering matrix and $S$ is the diagonal matrix of standard deviations of the columns of $G$.
Then $\widetilde{X}_j^T \widetilde{X} = R_{j,.} = R_{.,j}^T$, $\widetilde{X}_j^T \widetilde{X}_j = 1$ and
\[
u_j^{(t)} = \beta_j^{(t-1)} + \widehat{\beta}_j - R_{.,j}^T \beta^{(t-1)} ~,
\]
where $r_j = \widehat{\beta}_j =  \dfrac{\widehat{\gamma}_j}{\sqrt{n_j ~ \text{se}(\widehat{\gamma}_j)^2 + \widehat{\gamma}_j^2}}$ and $\widehat{\gamma}_j$ is the GWAS effect of variant $j$ and $n$ is the GWAS sample size \cite[]{mak2017polygenic,prive2021high}.
Then the most time-consuming part is computing $R_{.,j}^T \beta^{(t-1)}$.
To make this faster, instead of computing $R_{.,j}^T \beta^{(t-1)}$ at each iteration ($j$ and $t$), we can start with a vector with only 0s initially (for all $j$) since $\beta^{(0)} \equiv 0$, and then updating this vector when $\beta_j^{(t)} \neq \beta_j^{(t-1)}$ only. Note that only positions $k$ for which $R_{k,j} \neq 0$ must be updated in this vector $R_{.,j}^T \beta^{(t-1)}$.

In this new implementation of the lassosum model, the input parameters are the correlation matrix $R$, the GWAS summary statistics ($\widehat{\gamma}_j$, $\text{se}(\widehat{\gamma}_j)$ and $n_j$), and the two hyper-parameters $\lambda$ and $\delta$. 
Therefore, except for the hyper-parameters, this is the exact same input as for LDpred2 \cite[]{prive2020ldpred2}.
We try $\delta \in \{0.001, 0.005, 0.02, 0.1, 0.6, 3\}$ by default in lassosum2, instead of $s \in \{0.2, 0.5, 0.8, 1.0\}$ in lassosum.
For $\lambda$, the default in lassosum uses a sequence of 20 values equally spaced on a log scale between 0.1 and 0.001; we now use a sequence between $\lambda_0$ and $\lambda_0 / 100$ by default in lassosum2, where $\lambda_0 = \max_j \left|\widehat{\beta}_j\right|$ is the minimum $\lambda$ for which no variable enters the model because the L1-regularization is too strong.
Note that we do not provide an ``auto'' version using pseudo-validation (as in \cite{mak2017polygenic}) as we have not found it to be very robust (Figure \ref{fig:pseudoval} [TODO: REGENERATE THIS FIG AND CODE]).
Also note that, as in LDpred2, we run lassosum2 genome-wide using a sparse correlation matrix which assumes that variants further away than 3 cM are not correlated, and therefore we do not require splitting the genome into independent LD blocks anymore (as is required in lassosum).

>>>>

We design simulations where variants have different GWAS sample sizes, which is often the case when meta-analyzing GWAS from multiple cohorts without the same genome coverage.
Using 40,000 variants from chromosome 22 (Methods), we simulate quantitative phenotypes assuming an heritability of 20\% and 2000 causal variants.
For half of the variants, we use 100\% of 300,000 individuals for GWAS, but use only 80\% for one quarter of them and 60\% for the remaining quarter. 
We then run C+T, LDpred2-inf, LDpred2(-grid), LDpred2-auto, lassosum, lassosum2, PRS-CS and SBayesR by using either the true per-variant GWAS sample size, or the total sample size. 
We do not show results for SBayesR because it always diverged and for PRS-CS because the overlap with the LD reference they provide is too small.
Averaged over 10 simulations, when providing true per-variant GWAS sample sizes, squared correlations (in \%) between the polygenic scores and the simulated phenotypes are of [TODO: COMPLETE] (Figure \ref{fig:simu-misN}).
Note that C+T does not use this sample size information.
For methods that use the per-variant GWAS sample sizes, predictive performance slightly decreases [TODO: REPLACE NUMBERS] from 17.5 to 17.4 for lassosum, from 18.0 to 17.5 for lassosum2, but dramatically decreases for LDpred2 with new values of 11.5 for LDpred2-grid, 5.7 for LDpred2-auto and 6.8 for LDpred2-inf (Figure \ref{fig:simu-misN}).
Therefore, this extreme simulation scenario shows that LDpred2 can be sensitive to GWAS sample size misspecification, whereas lassosum (and lassosum2) seems little affected by this.

@@

\subsection*{GWAS using expected genotypes from imputation}

When using the expected genotypes from imputation (dosages), their standard deviations are usually lower than the expected value under Hardy-Weinberg equilibrium (cf.\ Methods section ``Standard deviations and imputation measures'').
In simulations (cf.\ Methods section ``Data for simulations''), we show that $\text{sd}(G_j)^{true} \approx \text{sd}(G_j)^{imp} / \sqrt{\text{INFO}_j}$, $\text{se}(\hat{\beta}_j)^{true} \approx \text{se}(\hat{\beta}_j)^{imp} \cdot \sqrt{\text{INFO}_j}$, and $\hat{\beta}_j^{true} \approx \hat{\beta}_j^{imp} \cdot \sqrt{\text{INFO}_j}$ (Figures \ref{fig:compare-sd}, \ref{fig:compare-beta}, and \ref{fig:compare-beta-se}).
This is the first correction of summary statistics we try in the simulations below. 
Note that we recompute INFO scores for the subset of 362,307 European individuals used in this paper as they sometimes differ from the ones reported by the UK Biobank (Figure \ref{fig:compare-info}).
Instead of using dosages to perform the GWAS, it has been argued that using multiple imputation (MI) would be more appropriate \cite[]{Palmer2016}.
In simulations, we show that $Z_j^{MI} \approx Z_j^{imp} \cdot \text{INFO}_j$ and $\hat{\beta}_j^{MI} \approx \hat{\beta}_j^{imp} \cdot \text{INFO}_j$ (Figure \ref{fig:compare-mi}).
This is the second correction of summary statistics we try in the simulations below, along with $n_j = N \cdot \text{INFO}_j$.

Using 40,000 variants from chromosome 22, we simulate quantitative phenotypes assuming an heritability of 20\% and 2000 causal variants using the ``true'' dataset (cf.\ Methods section ``Data for simulations'').
We run GWAS on the dosage dataset and use these summary statistics to run LDpred2 and lassosum2 with either no correction of the summary statistics, or with one of the two described above.
The LD reference used by LDpred2 and lassosum2 is computed from the validation set using the ``true'' dataset.

@@

\subsection*{GWAS sample size imputation}

In this paper, we extensively use the following formula
\begin{equation}\label{eq:approx-sd-lin}
\text{sd}(G_j) \approx \dfrac{\text{sd}(y)}{\sqrt{n_j ~ \text{se}(\hat{\beta}_j)^2 + \hat{\beta}_j^2}} ~,
\end{equation}
where $\hat{\beta}_j$ is the marginal (GWAS) effect of variant $j$, $n_j$ is the GWAS sample size associated with variant $j$, $y$ is the vector of phenotypes and ${G}_j$ is the vector of genotypes for variant $j$.
This formula is used in LDpred2 \cite{prive2020ldpred2,prive2021high}. Note that, for a binary trait for which logistic regression is used, we have instead
\begin{equation}\label{eq:approx-sd-log}
\text{sd}(G_j) \approx \dfrac{2}{\sqrt{n_j^\text{eff} ~ \text{se}(\hat{\beta}_j)^2 + \hat{\beta}_j^2}} ~,
\end{equation}
where $n_j^\text{eff} = \frac{4}{1 / n_j^\text{cases} + 1 / n_j^\text{controls}}$.

We can then impute $n_j$ from equation \eqref{eq:approx-sd-lin} using
\begin{equation}\label{eq:imputeN}
n_j \approx \dfrac{\text{var}(y) / \text{var}(G_j) - \hat{\beta}_j^2}{\text{se}(\hat{\beta}_j)^2} ~.
\end{equation}
In practice, we also bound this estimate to be between $0.5 \cdot N$ and $1.1 \cdot N$, where $N$ is the global sample size.

\subsection*{Standard deviations and imputation measures}

The MACH $\hat{r}_j^2$ measure for variant $j$ is given by $\hat{r}_j^2 = \frac{var(G_j)}{2 \hat{\theta} (1 - \hat{\theta})}$, where $\hat{\theta}$ is the estimated allele frequency \cite[]{marchini2010genotype}.
This measure is also highly concordant with the IMPUTE INFO measure \cite[]{marchini2010genotype}.
Therefore $\text{INFO}_j \approx \frac{var(G_j)}{2 \hat{\theta} (1 - \hat{\theta})}$.

@@

Instead of using a regularized version of the correlation matrix $R$ parameterized by $s$, $R_s = (1 - s) R + s I$, we use $R_{\delta} = R + \delta I$, which makes it clearer that lassosum is also using L2-regularization (therefore elastic-net). 
Then, from \cite{mak2017polygenic}, the solution from lassosum can be obtained by iteratively updating 
\[
\beta_j^{(t)} =
\begin{cases}
\text{sign}\left(u_j^{(t)}\right) \left(\left|u_j^{(t)}\right| - \lambda\right) / \left(\widetilde{X}_j^T \widetilde{X}_j + \delta\right) & \text{if } \left|u_j^{(t)}\right| > \lambda ~, \\
0 & \text{otherwise.}
\end{cases}
\]
where 
\[
u_j^{(t)} = r_j - \widetilde{X}_j^T \left( \widetilde{X} \beta^{(t-1)} - \widetilde{X}_j \beta_j^{(t-1)} \right) ~.
\]
Following notations from \cite{prive2020ldpred2} and denoting $\widetilde{X} = \frac{1}{\sqrt{n-1}} C_n G S^{-1}$, where $G$ is the genotype matrix, $C_n$ is the centering matrix and $S$ is the diagonal matrix of standard deviations of the columns of $G$.
Then $\widetilde{X}_j^T \widetilde{X} = R_{j,.} = R_{.,j}^T$, $\widetilde{X}_j^T \widetilde{X}_j = 1$ and
\[
u_j^{(t)} = \beta_j^{(t-1)} + \widehat{\beta}_j - R_{.,j}^T \beta^{(t-1)} ~,
\]
where $r_j = \widehat{\beta}_j =  \dfrac{\widehat{\beta}_j}{\sqrt{n_j ~ \text{se}(\widehat{\beta}_j)^2 + \widehat{\beta}_j^2}}$ and $\widehat{\beta}_j$ is the GWAS effect of variant $j$ and $n$ is the GWAS sample size \cite[]{mak2017polygenic,prive2021high}.
Then the most time-consuming part is computing $R_{.,j}^T \beta^{(t-1)}$.
To make this faster, instead of computing $R_{.,j}^T \beta^{(t-1)}$ at each iteration ($j$ and $t$), we can start with a vector with only 0s initially (for all $j$) since $\beta^{(0)} \equiv 0$, and then updating this vector when $\beta_j^{(t)} \neq \beta_j^{(t-1)}$ only. Note that only positions $k$ for which $R_{k,j} \neq 0$ must be updated in this vector $R_{.,j}^T \beta^{(t-1)}$.

In this new implementation of the lassosum model, the input parameters are the correlation matrix $R$, the GWAS summary statistics ($\widehat{\beta}_j$, $\text{se}(\widehat{\beta}_j)$ and $n_j$), and the two hyper-parameters $\lambda$ and $\delta$. 
Therefore, except for the hyper-parameters, this is the exact same input as for LDpred2 \cite[]{prive2020ldpred2}.
We try $\delta \in \{0.001, 0.005, 0.02, 0.1, 0.6, 3\}$ by default in lassosum2, instead of $s \in \{0.2, 0.5, 0.8, 1.0\}$ in lassosum.
For $\lambda$, the default in lassosum uses a sequence of 20 values equally spaced on a log scale between 0.1 and 0.001; we now use a sequence between $\lambda_0$ and $\lambda_0 / 100$ by default in lassosum2, where $\lambda_0 = \max_j \left|\widehat{\beta}_j\right|$ is the minimum $\lambda$ for which no variable enters the model because the L1-regularization is too strong.
Note that we do not provide an ``auto'' version using pseudo-validation (as in \cite{mak2017polygenic}) as we have not found it to be very robust (Figure \ref{fig:pseudoval}).
Also note that, as in LDpred2, we run lassosum2 genome-wide using a sparse correlation matrix which assumes that variants further away than 3 cM are not correlated, and therefore we do not require splitting the genome into independent LD blocks anymore (as is required in lassosum).
